Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-01-27 manuscript.tex
section 2, sans figure legend.

embryo cells marked by the expression of anterior-posterior marker TF (REF): from anterior to posterior SIX3, EMX2,
FOXG1, BARHL1, EN1 and HOXB3 (\textbf{Fig. 1g}).\par

@@

\subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}

@@

\subsection*{
  DeepNeuralTube: a sequence-to-function model revealing the enhancer code underlying early neuronal and neural crest 
  development
}

>>>>

embryo (REF): from anterior to posterior expression of SIX3, EMX2, FOXG1, BARHL1, EN1 and HOXB3 (\textbf{Fig. 1g}).\par

@@

\subsection*{Figure 1. Human neural tube organoids recapitulate early neuronal and neural crest development}

@@

\subsection*{
  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neuronal and neural crest 
  development
}

To gain insight into the sequence characteristics underlying cell type-specific activity of neural tube, neural crest
and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube_{organoid}$" and "$DeepNeuralTube_{embryo}$". The models use DNA sequence, of genomic regions, as input and
classify the regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class
fashion; \textbf(Fig. 2a)). The classes of the models are defined based on topic modeling. To get a fine-grained
definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest
(incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both the organoid cells (to
train $DeepNeuralTube_{organoid}$) and embryo cells (to train $DeepNeuralTube_{embryo}$). To validate the ability of the models to
predict cell type-specific enhancer \textit{activity}, we tested their performance on enhancers from the VISTA enhancer
browser (\textbf{REF}). Relevant for this study, this database contains enhancers active in the neural tube (n=349) and
facial mesenchyme (n=143). Neural tube enhancers could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{FIGURE SUPPLEMENT})), facial mesenchyme
enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$
(\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{FIGURE SUPPLEMENT})). Of note, a fraction of neural tube
enhancers (x.x \%) are predicted to be specific to pre-migratory neural crest by both models.\par
Next, we wanted to assess the explainability of the models. Concretely we wanted to test whether the model learned known
TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic and
calculated gradients, for each model, on the top 1,000 regions (as predicted by the model) for each topic followed by
TF-MoDISco (\textbf{REF}) to find \textit{de novo} motifs used by the models. For both known motifs and \textit{de novo}
motifs we calculated pairwise similarities using TOMTOM (\textbf{REF}) followed by clustering, resulting in 56 motif
clusters (\textbf{Fig. 2d}). Of note, most clusters contain known-motifs enriched in both organoid regions and embryo
(\textbf{Fig. 2e}) regions with the exception of cluster 13 which is organoid specific (AP1 motif; \textbf{SUPPLEMENTAL TABLE}) and cluster
26 which is embryo specific (CG-rich motif; \textbf{SUPPLEMENTAL TABLE}). Most clusters (\textbf{Fig. 2f-g} and
\textbf{SUPPLEMENTAL FIGURE}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.
Thus, $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par
As an illustrative example, we investigated an enhancer located in one of the introns of GLI3 (hs111.0; \textbf{SUPPLEMENTAL
FIGURE}). This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). Both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ explain this region similarly (Pearson correlation of x) and
highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). A version of this enhancer where 5\% of the
nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both
models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig. 2h}).\par
Finally, to also investigate the performance of the models on migratory neural crest, a cell type for which there are no
validated enhancers in the Vista enhancer browser, we selected a genomic region that is specifically accessible in the
migratory neural crest cell type of the organoid (\textbf{SUPPLEMENTAL FIGURE}). Both models predict this region
to be specific to the migratory neural crest class (\textbf{SUPPLEMENTAL FIGURE}) and nucleotide importance scores
reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). We tested the enhancer activity of this region
using a reporter assay in chicken embryos. Indeed the enhancer was specifically active in migratory neural crest 24
hours after injection in an HH4 embryo (\textbf{Fig. 2i}).
